Distance Sampling in \proglang{R}

### 2015-11-28 paper.Rmd
ER finishes first run through of edits
some questions included for DLM included in italics in .rmd

---
author:
  - name: David L Miller
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: dave@ninepointeightone.net
    url: http://converged.yt
  - name: Jeffrey L Laake
    affiliation: National Marine Mammal Laboratory
    address: >
      Alaska Fisheries Science Center
      7600 Sand Point Way N.E., Seattle, WA 98115, USA
    email: Jeff.Laake@noaa.gov
title:
  formatted: "Distance Sampling in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Distance Sampling in R"
  # For running headers, if needed
  #short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [distance sampling, abundance estimation, line transects, point transects, "\\proglang{R}"]
  plain:     [distance sampling, abundance estimation, line transects, point transects, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{booktabs}
output: rticles::jss_article
bibliography: jstatsoft.bib
---

@@

Distance sampling [@Buckland:2001vm; @Buckland:2004ts] encompases a suite of field methods and statistical models used to estimate the abundance of biological populations. Distance sampling field procedure can be thought of as an extension of plot sampling, where we wish to take into account the decreasing probability of detecting objects at increasing distance from the sampler. We do this by building a model for detectability and use quantities calculated from the detection function to adjust the observed counts to obtain an estimate of abundance.

@@

For both points and lines, once the geometry of the sampler (see "Detection functions") has been taken into account, the histogram of distances should show a decreasing number of observations with increasing distance from the sampler. For line transects we expect objects to be distributed unformly with respect to distance from the line and what makes our histogram decrease is the detectability. For point transects, we recognise that as distance from the point increases, the area of the circle encompassed increases with distance squared; hence the number of objects available to be detected is a linearly increasing function of distance from the point.

Using this histogram we can crudely estimate the drop-off in detectability by eye by tracing a line that approximates the tops of the histogram bars -- this is the detection function. \pkg{Distance} estimates the parameters for a fixed-form detection function using maximum likelihood estimation. We address possible models in detail below.

Figure \ref{fig:pointslines} shows examples of 500 individuals sampled using line and point transects (left column) and their corresponding histograms (right column).

@@

We demonstrate \pkg{Distance} using two data sets: one line transect and one point transect. These data sets have been chosen to be representative of the kind of data seen in practice.

@@

The line transect data is simulated data based on a survey of Antarctic minke whales (*Balaenoptera bonaerensis*). The data is simulated from models fitted to data from the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua]. They consist of 99 observations and include information on the effort expended and whether observations were in one of two geographical strata (near or far from land).

@@

The point transect data set consists of 1485 observations of Amakihi (*Hemignathus virens*; a Hawaiian songbird), collected at 41 points between 1992 and 1995. The data include distances and three covariates collected during the survey: observer ID (a three level factor), minutes after sunrise (continuous) and hours after sunrise (a six level factor). Data are analysed comprehensively in @Marques:2007ey.

@@

* In the simplest case, where one would simply like to estimate a detection function, all that is needed is a vector of distances.
* To include additional covariates into the detection function (see "Detection functions") we need to use a \code{data.frame}. The data.frame contains a column called \code{distance} and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}). Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether  an observation was detected (see "Extensions") and the columns described in the next bullet.
* If one would like to estimate abundance beyond area covered by expended survey effort, additional information is required. This consists of information about which transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (this may have any name and may be from pre- or post-survey stratification, see "Estimating abundance and variance") and that stratum's area (which has the same name as the stratum column, appended with \code{.Area}). (We refer to this data format as "flatfile" where all information is contained in one table.)

As we will see in "Extensions", further information is also required when we start using more complex models.

@@

We'll explore what the effects of including effort data are during analysis below.

@@

As mentioned above, the detection function describes the relationship between observed distances and probability of detection. The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (either from a line or point) and $\boldsymbol{\theta}$ is a vector of parmeters to be estimated. Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over the distances), so we must integrate out distance ($y$) from the detection function:
$$
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
$$
where $\pi(y)$ describes the distribution of objects with respect to the sampler and $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3] (letting $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point).

It is important that the detection function accurately models detectability near zero distance. We are less worried by its behaviour further away from 0. To ensure that the model is not overly influence by distances far from zero we truncate the distances beyond a given distance $w$ (known as the *truncation distance*). Fitting to distances at great distances from zero does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, 103-108, 151-153].

@@

* *Shoulder*: we expect observers to be able to see objects near them, not just those directly infront of them. For this reason, we expect the detection function to be "flat" near zero distance.
* *Non-increasing*: we don't think that observers should be more likely to see things further away than those nearer to them. This usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
* *Model robust*: models should be flexible enough to make many different shapes.
* *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these. We would like our models to be robust to us not including these factors.
* *Estimator efficiency*: we would like our models to have low variances, but only given they satisfy the other properties above (which, if they are satisfied, would give low bias).

Given these criteria, we can start to think about models for $g$.

@@

There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi]. \code{Distance} includes the most popular of these models. Here we'll detail the most popular detection function approach: "key function plus adjustments" (K+A).

@@

A disadvantage of K+A models is that we must resort to constrained optimisation (via the \pkg{Rsolnp} package) in order to ensure that the resulting detection function is monotonic non-increasing over the whole range.

@@

## half-normal with cosine
g <- function(x) exp(-x^2/(2*0.01^2))*(1+0.5*cos((2*pi*x)/0.025))
f <- function(x) g(x)/g(0)
curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with 1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
title(xlab="Distance", line=2)
title(ylab="Detection probability", line=2)
# 2 cosines
g <- function(x) exp(-x^2/(2*0.01^2))*(1-0.06*cos((2*pi*x)/0.025)+0.25*cos((3*pi*x)/0.025))
f <- function(x) g(x)/g(0)
curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with 2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
title(xlab="Distance", line=2)
title(ylab="Detection probability", line=2)

## hazard-rate with cosine
g <- function(x) (1 - exp(-(x/0.005)^-1.1))*(1-0.05*cos((2*pi*x)/0.025))
f <- function(x) g(x)/g(0)
curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with 1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
title(xlab="Distance", line=2)
title(ylab="Detection probability", line=2)
# 2 cosines
g <- function(x) (1 - exp(-(x/0.005)^-1.9))*
                  (1+0.05*cos((2*pi*x)/0.025) + 0.25*cos((3*pi*x)/0.025))
f <- function(x) g(x)/g(0)
curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with 2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
title(xlab="Distance", line=2)
title(ylab="Detection probability", line=2)
```

@@

There are many factors that can affect the probability of detecting an object. These include things like the observer, the vessel or platform used, the sea state, weather conditions and time of day (to name but a few). We assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).

Covariates can be included in this formulation by considering the scale parameter from the half-normal or hazard-rate detection functions as a linear model (on the exponential scale) of the ($J$) covariates ($\mathbf{z}$; a vector of length $J$ for each observation):
$$
\sigma(\mathbf{z}) = \exp(\beta_0 + \sum_{j=1}^J \beta_j z_j).
$$
In the next section we'll discuss model selection.

Including covariates has an important implication for our calculation of detectability. Since we don't know what the true distribution of the covariates is we must calculate the probability of detection conditional on the observed values of the covariates:
$$
p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
$$
where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$. So for covariate models, we are calculating a value of "average" probability of detection (again the average is in the sense of distance) per observation. There will be as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.

Another important consideration is that K+A models which include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations, as we don't have any model for the distribution of the covariates. For this reason, we advise against using both adjustments and covariates in a detection function [see @Miller:2015hw for an example of when this can be problematic].

@@

The workhorse of detection function fitting in \pkg{Distance} is the \code{ds} function. Here we show off the formulations for the detection function that we've seen above for both the minke whale and amakihi data.

@@

We can fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
```{r minke-hn}
minke_hn <- ds(minke, truncation=1.5)
```
Note that \code{ds} will automatically select adjustment terms by AIC and shows it's selection steps as it goes.

Figure \ref{fig:minkeamakihi} (top left plot) shows the result of calling \code{plot} on the resulting model object. We can also call \code{summary} on the model object to get summary information about the fitted model (though we postpone this until the next section).

We can specify the form of the detection function via the \code{key=} argument to \code{ds}. For example, a hazard rate model can be fitted as:
```{r minke-hr-cos}
minke_hrcos <- ds(minke, truncation=1.5, key="hr")
```
Again, \code{ds} fits a model with cosine adjustments (the default) but finds the AIC improvement to be insufficient to select the adjustment.

@@

By default \code{ds} assumes that the data given to it is line transect, but we can switch to points using \code{transect="point"}. Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate can be specified by [truncating at 82.5m, given in @Marques:2007ey]:
```{r amakihi-hr-obs}
amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
                     key="hr", formula=~obs)
```
As with the minke whale model, we can plot the resulting detection function (Figure \ref{fig:minkeamakihi}, centre column). Since for the amakihi we used covariates in the detection function, the plot will show the detection function averaged over levels/values of the covariate. Points on the plot indicate probability of detection  for each observation. For the `amakihi_hr_obs` model we see fairly clear levels of the observer covariate in the points. Looking at the left panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear when adding minutes after sunrise as a covariate into the model:
```{r amakihi-hr-obs-mas, warning=FALSE}
amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
                         key="hr", formula=~obs+mas)
```

The bottom row of Figure \ref{fig:minkeamakihi} shows the probability density functions for the amakihi models. These often are easier to interpret than the detection functions, as for point data we must rescale the histogram when plotting the detection function to take into account the geometry of the point sampler.

@@

As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give us some useful information about our fitted model. For example for our hazard-rate model for the amakihi, with observer as a covariate:
```{r amakihi-summary}
summary(amakihi_hr_obs)
```
This summary information includes various details of the data and model specification, as well as the values of the coefficents ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and it's uncertainty. The final line gives an estimate of abundance for the area covered by the survey (again, this will be covered in detail in the next section).

@@

To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot). In our case the CDF evaluates the probability of observing an object at a distance less than or equal to some given value. The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance. In essence we're asking "is the number of observations up to a given distance in line with what the model says they should be?" (where 'given values' are the observed distances). As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.

We can inspect Q-Q plots visually, though for a large number of models this can be tiresome and prone to subjective judgements. Instead we can quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cramer-von Mises test [@Burnham:2004vd]. Both test to see if the points from the EDF and CDF are from the same distribution. The Kolmogorov-Smirnov uses the test statistic of the largest difference between a point on the Q-Q plot and the line $y=x$, whereas the Cramer-von Mises test uses the sum of all the differences. As it takes into account more information and is therefore more powerful, the Cramer-von Mises is generally preferred. A significant result from either tests indicates that the EDF and CDF do not come from the same distribution (and therefore the model is does not fit the data well).

We can generate a Q-Q plot and test results using the \code{gof_ddf} function. We can see the goodness of fit tests below for two models for the amakihi data, first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments), then calculating goodness of fit for that model and our hazard-rate model with observer and minutes after sunrise included:
```{r amakihi-gof, fig.keep="none"}
amakihi_hr <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
gof_ds(amakihi_hr)
gof_ds(amakihi_hr_obs_mas)
```
The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}.

```{r, amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data.\\label{amakihi-qq}", echo=FALSE, results="hide"}
par(mfrow=c(1,2))
gof_ds(amakihi_hr, asp=1)
gof_ds(amakihi_hr_obs_mas, asp=1)
```

@@

Though fiting the detection function is the primary modelling step in distance sampling, we are really interested in estimating detectability, and from that abundance. We also wish to calculate our certainty in each abundance estimate. This section addresses these issues mathematically before showing how to estimate abundance and its variance in R.

@@

First, in order to estimate abundance in the covered area ($\hat{N}_\text{C}$), we use the estimates of detection probability (the $\{p_i; i=1,\ldots,n\}$, above) in a Horvitz-Thompson like estimator:
\begin{equation}
\hat{N}_\text{C} = \sum_{i=1}^n\frac{s_i}{\hat{p}_i},
\label{ht}
\end{equation}
where $s_i$ are the sizes of the observed groups of objects, which is equal to 1 if objects only occur singly [@Borchers:2004wr]. @thompson2012sampling is the cannonical reference to this type of estimator; intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$) to account for incomplete detection -- we then sum to obtain the abundance estimate. For models that do not include covariates, $\hat{p}_i$ is equal for all $i$, so this is equivalent to summing the groups and inflating that sum by the corresponding $\hat{p} (=\hat{p}_i \forall i)$.

@@

We can use the Horvitz-Thompson-like estimator to calculate the "average" detectability for models which include covariates. We can consider what single detectability value would give the estimated $\hat{N}$ and therefore calculate:
$$
\hat{P_a} = n/\hat{N}_\text{C}.
$$
This can be a useful summary statistic and is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.

@@

We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas *strata*. A stratum may be defined by, for example, a different habitat type or by the sex of the animal (or some combination) which may be of interest for biological or management reasons. In order to calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like sex, we would have the areas be the same but if we were using say forested vs. wetland habitat the areas of those strata would be different). Finally, we must also know which stratum a given transect lies in.

As an example the minke whale data consists of two strata: \code{North} and \code{South} relating to a stratum further away and nearer the Antarctic ice edge, respectively. Figure \ref{fig:minke-strata} shows the two strata, along with observations and transect lines.

>>>>

---
author:
  - name: David L Miller
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: dave@ninepointeightone.net
    url: http://converged.yt
  - name: Eric Rexstad
    affiliation: University of St Andrews
    address: >
      Centre for Research into Ecological and Environmental Modelling,
      The Observatory,
      St Andrews, Fife KY16 9LZ, Scotland
    email: Eric.Rexstad@st-andrews.ac.uk
  - name: Jeffrey L Laake
    affiliation: National Marine Mammal Laboratory
    address: >
      Alaska Fisheries Science Center
      7600 Sand Point Way N.E., Seattle, WA 98115, USA
    email: Jeff.Laake@noaa.gov
title:
  formatted: "Distance Sampling in \\proglang{R}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Distance Sampling in R"
  # For running headers, if needed
  #short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [distance sampling, abundance estimation, line transects, point transects, "\\proglang{R}"]
  plain:     [distance sampling, abundance estimation, line transects, point transects, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{booktabs}
output: rticles::jss_article
bibliography: jstatsoft.bib
---

@@

Distance sampling [@Buckland:2001vm; @Buckland:2004ts] encompasses a suite of field methods and statistical models used to estimate the abundance of biological populations. Distance sampling field procedure can be thought of as an extension of plot sampling, where we wish to take into account the decreasing probability of detecting objects at increasing distance from the sampler. We do this by building a model for detectability and use quantities calculated from the detection function to adjust the observed counts to obtain an estimate of abundance.

@@

For both points and lines, given the geometry of the sampler (see "Detection functions"), the histogram of distances should show a decreasing number of observations with increasing distance from the sampler. For line transects we expect objects to be distributed uniformly with respect to distance from the line with detectability causing the number of detections to decrease. For point transects, we recognise that as distance from the point increases, the area of the circle encompassed increases with distance squared; hence the number of objects available to be detected is a linearly increasing function of distance from the point.

Using this histogram we can crudely estimate the drop-off in detectability by eye by tracing a line that approximates the tops of the histogram bars -- this is the detection function. \pkg{Distance} estimates the parameters for a fixed-form detection function using maximum likelihood estimation. We address possible models below.

Figure \ref{fig:pointslines} shows sampling of a population of 500 individuals using line and point transects (left column) and their corresponding histograms (right column).

@@

We demonstrate \pkg{Distance} using two data sets: one line transect and one point transect. These data sets have been chosen to be representative of data seen in practice.

@@

The line transect data is based on a survey of Antarctic minke whales (*Balaenoptera bonaerensis*). The data are simulated from models fitted to data from the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua]. They consist of 99 observations and include geographical strata (near or distant from ice edge) and transect lengths on the effort expended.

@@

The point transect data set consists of 1485 observations of Amakihi (*Hemignathus virens*; a Hawaiian songbird), collected at 41 points between 1992 and 1995. The data include distances and two covariates collected during the survey: observer (a three level factor), time after sunrise (transformed to minutes (continuous) or hours (factor) covariates). Data are analysed comprehensively in @Marques:2007ey.

@@

* In the simplest case, where the objective is to estimate a detection function, all that is needed is a vector of distances.
* To include additional covariates into the detection function (see "Detection functions") we use a \code{data.frame}. The data.frame contains a column called \code{distance} and additional named columns for covariates potentially useful in modelling detectability (for example \code{observer} or \code{seastate}). Some column names are reserved: \code{object} for an observation identifier (see "Extensions"), \code{size} for group or cluster size (see "Detection functions" and "Abundance and variance estimation"), \code{detected} for whether an observation was detected (see "Extensions") and the columns described in the next bullet.
* To estimate abundance beyond area covered by expended survey effort, additional information is required. This consists of information about which transect and occasion the observation was made (the \code{Sample.Label}), a column named \code{Effort} which gives the effort associated with that sample (for lines their length and for points the number of times that point was visited), the stratum the sample was located in (this may have any name and may be from pre- or post-survey stratification, see "Estimating abundance and variance") and strata areas (which has the same name as the stratum column, appended with \code{.Area}). We refer to this data format as "flatfile" where all information is contained in one table.

As we will see in "Extensions", further information is also required when using more complex models.

@@

We will explore below the effects of including effort data during analysis.

@@

The detection function describes the relationship between observed distances and probability of detection. The detection function itself models the probability $\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)$ and is usually denoted $g(y; \boldsymbol{\theta})$ where $y$ is distance (either from a line or point) and $\boldsymbol{\theta}$ is a vector of parameters to be estimated. Our goal is to estimate an *average probability of detection* ($p$, average in the sense of an average over distances), so we must integrate out distance ($y$) from the detection function:
$$
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
$$
where $\pi(y)$ describes the distribution of objects with respect to the sampler and $\pi(x)=1/w$ for line transects and $\pi(r)=\frac{2r}{w^2}$ for point transects, taking into account the geometry of the sampler [@Buckland:2001vm, Chapter 3] (letting $x$ denote a perpendicular distance from a line and $r$ denote radial distance from a point).

It is crucial that the detection function accurately models detectability at small distances. We are less worried by its behaviour further away from 0. To ensure that the model is not overly influence by distances far from zero, we truncate the distances beyond a given distance $w$ (known as the *truncation distance*). Fitting to distances at great distances from zero does not demonstrably improve the precision of abundance estimates [@Buckland:2001vm, 103-108, 151-153].

@@

* *Shoulder*: we expect observers to be able to see objects near them, not just those directly adjacent to them. For this reason, we expect the detection function to be horizontal near zero distance.
* *Non-increasing*: we do not think that observers should be more likely to see things distant from the transect than those near the transect. If this occurs, it usually indicates an issue with field procedure (that the distribution of objects with respect to the line, $\pi(y)$ is not what we expect), so we do not want the detection function to model this.
* *Model robust*: models should be flexible enough to make many different shapes.
* *Pooling robust*: many factors can affect the probability of detection and it is not possible to measure all of these. We would like models to produce unbiased results without inclusion of these factors.
* *Estimator efficiency*: we would like models to have low variances, but only given they satisfy the other properties above (which, if they are satisfied, would give low bias).

Given these criteria, we can formulate models for $g$.

@@

There is a wide literature on possible formulations for the detection function [@Buckland:1992fa; @Eidous:2005bj; @Becker:2009cj; @Giammarino:2014eg; @Miller:2015hw; @Becker:2015fi]. \code{Distance} includes the most popular of these models. Here we will detail the most popular detection function approach: "key function plus adjustments" (K+A).

@@

A disadvantage of K+A models is that we must resort to constrained optimisation (via the \pkg{Rsolnp} package) to ensure that the resulting detection function is monotonic non-increasing over the whole range.

@@

## half-normal with cosine
g <- function(x) exp(-x^2/(2*0.01^2))*(1+0.5*cos((2*pi*x)/0.025))
f <- function(x) g(x)/g(0)
curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with \n1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
title(xlab="Distance", line=2)
title(ylab="Detection probability", line=2)
# 2 cosines
g <- function(x) exp(-x^2/(2*0.01^2))*(1-0.06*cos((2*pi*x)/0.025)+0.25*cos((3*pi*x)/0.025))
f <- function(x) g(x)/g(0)
curve(f, from=0, to=0.025, xlab="", ylab="", main="Half-normal with \n2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
title(xlab="Distance", line=2)
title(ylab="Detection probability", line=2)

## hazard-rate with cosine
g <- function(x) (1 - exp(-(x/0.005)^-1.1))*(1-0.05*cos((2*pi*x)/0.025))
f <- function(x) g(x)/g(0)
curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n1 cosine adjustment",xlim=c(0,0.025), ylim=c(0,1))
title(xlab="Distance", line=2)
title(ylab="Detection probability", line=2)
# 2 cosines
g <- function(x) (1 - exp(-(x/0.005)^-1.9))*
                  (1+0.05*cos((2*pi*x)/0.025) + 0.25*cos((3*pi*x)/0.025))
f <- function(x) g(x)/g(0)
curve(f, from=0, to=0.025, xlab="", ylab="", main="Hazard-rate with \n2 cosine adjustments",xlim=c(0,0.025), ylim=c(0,1))
title(xlab="Distance", line=2)
title(ylab="Detection probability", line=2)
```

@@

There are many factors that can affect the probability of detecting an object. These include things like the observer, the vessel or platform used, the sea state, weather conditions and time of day. We assume that these variables affect detection only via the scale of the detection function (and do not affect the shape).

Covariates can be included in this formulation by considering the scale parameter from the half-normal or hazard-rate detection functions as a linear model (on the exponential scale) of the ($J$) covariates ($\mathbf{z}$; a vector of length $J$ for each observation):
$$
\sigma(\mathbf{z}) = \exp(\beta_0 + \sum_{j=1}^J \beta_j z_j).
$$
In the next section we will discuss model selection.

Including covariates has an important implication for our calculation of detectability. Because we do not know what the true distribution of the covariates is we must calculate the probability of detection conditional on the observed values of the covariates:
$$
p(\mathbf{z_i}) = \int_0^w \pi(y) g(y, \mathbf{z_i}; \boldsymbol{\theta}) dy,
$$
where $\mathbf{z_i}$ is the vector of $J$ covariates associated with observation $i$. For covariate models, we are calculating a value of "average" probability of detection (average in the sense of distance) per observation. There will be as many unique values of $p(\mathbf{z_i})$ as there are unique covariate combinations in our data.

K+A models that include covariates and one or more adjustments cannot be guaranteed to be monotonic non-increasing for all covariate combinations, as we do not have any model for the distribution of the covariates. For this reason, we advise against using both adjustments and covariates in a detection function [see @Miller:2015hw for an example of when this can be problematic].

@@

The workhorse of detection function fitting in \pkg{Distance} is the \code{ds} function. Here we show off the formulations for the detection function we have seen above applied to the minke whale and amakihi data.

@@

We can fit a model to the minke whale data, setting the truncation at 1.5km and using the default options in \code{ds} very simply:
```{r minke-hn}
minke_hn <- ds(minke, truncation=1.5)
```
Note that \code{ds} will automatically select adjustment terms by AIC and shows the selection steps.

Figure \ref{fig:minkeamakihi} (top left plot) shows the result of calling \code{plot} on the resulting model object. We can also call \code{summary} on the model object to get summary information about the fitted model (we postpone this to the next section).

We can specify the form of the detection function via the \code{key=} argument to \code{ds}. For example, a hazard rate model can be fitted as:
```{r minke-hr-cos}
minke_hrcos <- ds(minke, truncation=1.5, key="hr")
```
\code{ds} fits a model with cosine adjustments (the default) but finds the AIC improvement to be insufficient to select the adjustment.

@@

By default \code{ds} assumes the data given to it is line transect, but we can switch to points using \code{transect="point"}. Including covariates in the scale is via the \code{formula=~...} argument to \code{ds}; a hazard-rate model for the amakihi that includes observer as a covariate can be specified by [truncating sightings at 82.5m, given in @Marques:2007ey]:
```{r amakihi-hr-obs}
amakihi_hr_obs <- ds(amakihi, truncation=82.5, transect="point",
                     key="hr", formula=~obs)
```
As with the minke whale model, we can plot the resulting detection function (Figure \ref{fig:minkeamakihi}, centre column). Because the amakihi model included covariates, the plot shows the detection function averaged over levels/values of the covariate. Points on the plot indicate probability of detection for each observation. For the `amakihi_hr_obs` model we see fairly clear levels of the observer covariate in the points. Looking at the left panel of Figure \ref{fig:minkeamakihi}, we can see this is less clear when adding minutes after sunrise as a covariate into the model:
```{r amakihi-hr-obs-mas, warning=FALSE}
amakihi_hr_obs_mas <- ds(amakihi, truncation=82.5, transect="point",
                         key="hr", formula=~obs+mas)
```

The bottom row of Figure \ref{fig:minkeamakihi} shows the probability density functions for the amakihi models. Probability density function plots give a better sense of fit of the model than the detection functions, as for point data we must rescale the histogram when plotting the detection function to take into account the geometry of the point sampler.

** ER suggests the g(x) plots for the amakihi be deleted **

@@

As with models fitted using \code{lm} or \code{glm} in \proglang{R}, we can use \code{summary} to give useful information about our fitted model. For example for our hazard-rate model for the amakihi, with observer as a covariate:
```{r amakihi-summary}
summary(amakihi_hr_obs)
```
This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an "average" value for the detectability (see "Estimating abundance and variance" for details on how this is calculated) and its uncertainty. The final line gives an estimate of abundance for the area covered by the survey (covered in the next section).

@@

To judge goodness of fit for detection functions when exact distances are used, we want to compare the cumulative distribution function (CDF) and empirical distribution function (EDF) for the detection function via a quantile-quantile plot (Q-Q plot). In our case the CDF evaluates the probability of observing an object at a distance less than or equal to some given value. The EDF gives the proportion of observations for which the CDF is less than or equal to that of a given distance. This can be interpreted as assessing whether the number of observations up to a given distance in line with what the model says they should be (where 'given values' are the observed distances). As usual for Q-Q plots, "good" models will have values close to the line $y=x$, poor models will show more deviations from that line.

We can inspect Q-Q plots visually, though this is prone to subjective judgments. Instead we can quantify the Q-Q plot's information using a Kolmogorov-Smirnov or Cramer-von Mises test [@Burnham:2004vd]. Both test whether points from the EDF and CDF are from the same distribution. The Kolmogorov-Smirnov uses the test statistic of the largest difference between a point on the Q-Q plot and the line $y=x$, whereas the Cramer-von Mises test uses the sum of all the differences. As it takes into account more information and is therefore more powerful, the Cramer-von Mises is generally preferred. A significant result from either tests indicates that the EDF and CDF do not come from the same distribution (and therefore the model is does not fit the data well).

We can generate a Q-Q plot and test results using the \code{gof_ddf} function. We can see the goodness of fit tests below for two models for the amakihi data (Figure \ref{amakihi-qq}), first fitting a half-normal model without covariates or adjustments (not setting \code{adjustment=NULL} will force \code{ds} to fit a model with no adjustments), then calculating goodness of fit for that model and our hazard-rate model with observer and minutes after sunrise included:
```{r amakihi-gof, fig.keep="none"}
amakihi_hr <- ds(amakihi, truncation=82.5, transect="point", key="hn", adjustment=NULL)
gof_ds(amakihi_hr)
gof_ds(amakihi_hr_obs_mas)
```
The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}.

```{r, amakihi-qq-comp, fig.width=6, fig.height=3, fig.cap="Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates) and hazard-rate model with observer and minutes after sunrise for the amakihi data.\\label{amakihi-qq}", echo=FALSE, results="hide"}
par(mfrow=c(1,2))
gof_ds(amakihi_hr, asp=1, cex=0.3)  # shrinking dot size to obliterate 45-deg line less
gof_ds(amakihi_hr_obs_mas, asp=1, cex=0.3)
```

@@

Though fitting the detection function is the primary modelling step in distance sampling, we are really interested in estimating detectability, and from that abundance. We also wish to calculate our certainty in each abundance estimate. This section addresses these issues mathematically before showing how to estimate abundance and its variance in \proglang{R}.

@@

First, to estimate abundance in the covered area ($\hat{N}_\text{C}$), we use the estimates of detection probability (the $\{p_i; i=1,\ldots,n\}$, above) in a Horvitz-Thompson like estimator:
\begin{equation}
\hat{N}_\text{C} = \sum_{i=1}^n\frac{s_i}{\hat{p}_i},
\label{ht}
\end{equation}
where $s_i$ are the sizes of the observed groups of objects, which is equal to 1 if objects only occur singly [@Borchers:2004wr]. @thompson2012sampling is the canonical reference to this type of estimator; intuitively, we can think of the estimates of detectability ($\hat{p}_i$) as "inflating" the group sizes ($s_i$) to account for incomplete detection -- we then sum over all $n$ detections to obtain the abundance estimate. For models that do not include covariates, $\hat{p}_i$ is equal for all $i$, so this is equivalent to summing the groups and inflating that sum by the corresponding $\hat{p} (=\hat{p}_i \forall i)$.

@@

*have L and K been previously defined?*

We can use the Horvitz-Thompson-like estimator to calculate the "average" detectability for models which include covariates. We can consider what single detectability value would give the estimated $\hat{N}$ and therefore calculate:
$$
\hat{P_a} = n/\hat{N}_\text{C}.
$$
This can be a useful summary statistic and is included in the \code{summary} output and the table produced by \code{summarize_ds_models}.  *anything else to say about this, seems to mess up folks, including me*

@@

We may wish to calculate abundance estimates for some sub-regions of the study region, we call these areas *strata*. For example, strata may be defined by habitat types or animal gender (or some combination) which may be of interest for biological or management reasons. To calculate estimates for a given stratification each observation must occur in a stratum which must be labelled with a \code{Region.Label} and have a corresponding \code{Area} (if we are using an animal characteristic like gender, we would have the areas be the same but if we were using say forested vs. wetland habitat the areas of those strata would be different). Finally, we must also know the stratum in which transects lie.

As an example the minke whale data consists of two strata: \code{North} and \code{South} relating to a stratum further away and nearer the Antarctic ice edge, respectively. Figure \ref{fig:minke-strata} shows the two strata, along with observations and transect lines.  *figure number messed up*
